### 2021-06-29 ch5langevin.tex
got some stuff figured out .

Bulk bed load transport rates show wide and frequent fluctuations which originate from coupling between the fluid and granular phases.
Due to these fluctuations, measured transport rates often show extremely slow convergence through time, and predicted rates often deviate from measured values by several orders of magnitude.
These challenges limit numerous ecological and engineering applications that rely on sediment transport predictions.
In recent decades, stochastic formulations of the bed load flux have become increasingly popular for their potential to predict the mean transport rates required by applications while also predicting fluctuations, quantifying the dependence of measurements on the observation scale, and linking bulk transport characteristics to the ``microscopic" dynamics of individual grains.
Recent indications that sediment transport fluctuations might explain longstanding and unsolved problems in alluvial channel stability, such as channel width maintenance \citep{Boltzman, Lajeunesse2020} and bedform initiation \citep{Ancey2014,Bohorquez2016} provide additional motivation for these stochastic formulations.
Many stochastic approaches express downstream transport rates as a sum over the instantaneous streamwise velocities of all particles in motion within a control volume.
In these approaches, the instantaneous velocity distribution of sediment particles therefore becomes an object of crucial importance.
Unfortunately, current understanding of this distribution remains limited, and no mechanistic models have yet been presented that describe the full range of experimental observations. Here, we take some steps toward rectifying these shortcomings.

High-speed video experiments have measured different streamwise particle velocity distributions without providing much understanding as to why one distribution or another appears.
One set of studies has shown exponential particle velocity distributions \citep{Charru2004,Lajeunesse2010,Roseberry2012,Seizilles2014,Fathel2015,Fathel2016}. 
These experiments involve uniformly-sized small sands or glass beads ($~0.05-2$mm) having typical Stokes numbers $St \sim 1-10$. Their flow conditions are generally sub-critical ($Fr<1$) and turbulent ($Re>5000$) but not always: flows in \citet{Lajeunesse2010} were super-critical ($Fr>1$), and flows in \citet{Charru2004} and \citet{Sezilles2014} were viscous.
A second set of studies have shown Gaussian particle velocity distributions \citep{Ancey2014,Heyman2016,Martin2012}. In these experiments, particles are typically larger ($2-8$mm) uniformly-sized gravels or glass beads having higher Stokes numbers ($St \sim 10-500$). In all Gaussian cases, flows are turbulent ($Re>5000$) and super-critical ($Fr>1$).
Two other experiments display velocity distributions that are intermediate between exponential and Gaussian and appear more like a Gamma distribution \citep{Houssais2012, Liu2019}. The \citet{Houssais2012} experiments involved a binomal distribution of glass beads with diameters $0.7$mm and $2.2$mm in turbulent and supercritical flow conditions. They resolved the velocity distributions for larger grains only. The \citet{Liu2019} experiments used uniformly-graded sand having median diameter $1.1$mm. Flows were again turbulent and subcritical.
From this experimental record, we can summarize that the shape of the velocity distribution does not consistentely relate to whether a flow is super or sub-critical ($Fr$), whether sediment grains are natural (sand, gravel) or synthetic (beads), or whether the flow is laminar or turbulent ($Re$). However, the typical Stokes numbers of particles do seem to increase monotonically from exponential velocity experiments (where $St \sim 1$), to intermediate (Gamma-like) experiments (where $St \sim 10$), and eventually to Gaussian experiments (where $St \sim 10^2$.) Apparently, the shape of streamwise bed load velocity distributions depends on the particle size.

@@

We developed a Langevin description of bed load sediment transport which includes episodic collisions between particles and the bed.
The model relates the shape of the instantaneous streamwise particle velocity distribution to the elasticity of particle-bed collisions,
generalizes earlier approaches available in the literature which did not treat episodic collisions \citep{Ancey2014,Fan2014}, and provides a new physical explanation for the different streamwise sediment velocity distributions resolved in experiments.
Although in reality, the turbulent forces on moving sediment particles vary in a complex spatio-temporal way, we have approximated the fluid forces on bed load particles as spatially uniform Gaussian white noise. Even though the non-Gaussian aspects of fluid turbulence certainly do impact sediment entrainment \citep{Coleman2019,Celik2014}, this flow model appears more or less justified since sediment transport experiments provide similar velocity distributions regardless of whether the flow is viscous or turbulent \citep{Lajeunesse2010, Charru2004}, and since particle relaxation times are relatively long compared to the timescales of turbulent fluctuations \citep{}.
We modelled particle-bed collision forces as a sequence of instantaneous impulses where the intervals between successive collisions were characterized as exponential random variables. The effect of each collision on the streamwise particle velocities was parameterized by a restitution-like coefficient.
Although such approximate descriptions of particle-particle collisions are common in the theory of granular gases, the setting here is somewhat different than grains in air. Because particles within a viscous flow interact at a distance, we should expect the collision model will become poor when the time between subsequent collisions becomes small. Therefore, although the model seems appropriate for saltation, it could become questionable for the ``reptation" transport mode when the times between subsequent particle-bed collisions are short. Of course, more realistic flow and collision forces could always be incorporated into Langevin equations for bed load transport. Unfortunately, numerical methods would likely be required to interpret the resulting equations, in contrast to the analytical approaches applied here.

Sediment transport experiments reveal correlations between particle size and the shape of the bed load velocity distribution.
Experiments with smaller particles tend to give exponential distributions, and those with larger particles give Gaussian distributions.
In fluid dynamics, the dissipation characteristics of particle-particle collisions in viscous flows are known to depend on the particle size and approach velocity through the Stokes number. 
In kinetic theory, it is known that gases of ideal elastic particles generate Gaussian (Boltzmann) velocity distributions, while gases of inelastic particles generate non-Gaussian distributions.
Taken together, these ideas suggest that we might relate the shape of the particle velocity distribution to particle size.
We can estimate typical Stokes numbers of colliding bed load particles in experiments as $...$, using with the flow shear velocity and mean streamwise sediment velocity to calculate $V$. Estimating in this way, transport experiments with exponential velocities have $St \sim 1-10$, those with neither exponential nor Gaussian velocities have $St \sim 10-100$, and those with Gaussian velocities have $St > 100$.
In experiments relating restitution coefficient to Stokes number for idealized collisions, restitution coefficients vary sharply from $0$ to $1$ as $St$ ranges from $1$ to $500$ \cite{Marshall2001,Joseph2001,Yang2006}.
Although collision geometry and grain shape certainly complicate the narrative, these values of $St$ are consistent with our model conclusion that the shape of the particle velocity depends on the elasticity of collisions.

@@

>>>>

Bulk bed load transport rates show wide and frequent fluctuations which originate from coupling between the fluid and granular phases.
Due to these fluctuations, measured transport rates often show extremely slow convergence through time, and predicted rates often deviate from measured values by several orders of magnitude.
These challenges limit numerous ecological and engineering applications that rely on sediment transport predictions.
In recent decades, stochastic formulations of the bed load flux have become increasingly popular for their potential to predict the mean transport rates required by applications while also predicting fluctuations, quantifying the dependence of measurements on the observation scale, and linking bulk transport characteristics to the ``microscopic" dynamics of individual grains.
Recent indications that sediment transport fluctuations might explain longstanding and unsolved problems in alluvial channel stability, such as channel width maintenance \citep{Boltzman, Lajeunesse2020} and bedform initiation \citep{Ancey2014,Bohorquez2016} provide additional motivation to develop these stochastic approaches.
One subset of these methods expresses downstream transport rates as a sum over the instantaneous streamwise velocities of all particles in motion within a control volume.
These approaches highlight the instantaneous velocity distribution of sediment particles as an object of crucial importance.
Unfortunately, current understanding of bedload velocity distributions remains limited.
We have as of yet no consensus on the shape of the bedload velocity distribution (is it Exponential, Gaussian, Gamma, or some other shape?), and no mechanistic models have yet been developed to describe the full range of experimental observations.
Here, we develop a stochastic model of particle velocities to addresses these issues.

Particle trajectories are a compromise between turbulent drag and particle-bed collision \citep{Wiberg1985}. We therefore anticipate that the Stokes number, which is known to characterize colliding bodies within a viscous fluid \citep{}, is an important dimensionless parameter of the bedload velocity distribution.
The Stokes number is defined as ... where ....

High-speed video experiments have measured different streamwise particle velocity distributions without providing much understanding as to why one distribution or another appears in a given set of hydraulic and sedimentary conditions.One set of studies has shown exponential particle velocity distributions \citep{Charru2004,Lajeunesse2010,Roseberry2012,Seizilles2014,Fathel2015,Fathel2016}.
These experiments involve uniformly-sized small sands or glass beads ($~0.05-2$mm) having typical Stokes numbers $St \sim 1-10$. Their flow conditions are generally sub-critical ($Fr<1$) and turbulent ($Re>5000$) but not always: flows in \citet{Lajeunesse2010} were super-critical ($Fr>1$), and flows in \citet{Charru2004} and \citet{Sezilles2014} were viscous.
A second set of studies have shown Gaussian particle velocity distributions \citep{Ancey2014,Heyman2016,Martin2012}. In these experiments, particles are typically larger ($2-8$mm) uniformly-sized gravels or glass beads having higher Stokes numbers ($St \sim 10-500$). In all Gaussian cases, flows are turbulent ($Re>5000$) and super-critical ($Fr>1$).
Two other experiments display velocity distributions that are intermediate between exponential and Gaussian and appear more like a Gamma distribution \citep{Houssais2012, Liu2019}. The \citet{Houssais2012} experiments involved a binomal distribution of glass beads with diameters $0.7$mm and $2.2$mm in turbulent and supercritical flow conditions. They resolved the velocity distributions for larger grains only. The \citet{Liu2019} experiments used uniformly-graded sand having median diameter $1.1$mm. Flows were again turbulent and subcritical.
From this experimental record, we can summarize that the shape of the velocity distribution does not consistentely relate to whether a flow is super or sub-critical ($Fr$), whether sediment grains are natural (sand, gravel) or synthetic (beads), or whether the flow is laminar or turbulent ($Re$). However, the typical Stokes numbers of particles do seem to increase monotonically from exponential velocity experiments (where $St \sim 1$), to intermediate (Gamma-like) experiments (where $St \sim 10$), and eventually to Gaussian experiments (where $St \sim 10^2$.) Apparently, the shape of streamwise bed load velocity distributions depends on the particle size.

@@

Here, I developed a Langevin description of bed load sediment transport which includes episodic collisions between particles and the bed.
The model relates the shape of the instantaneous streamwise particle velocity distribution to the elasticity of particle-bed collisions,
generalizes earlier approaches available in the literature which did not treat episodic collisions \citep{Ancey2014,Fan2014}, and provides a new physical explanation for the different streamwise sediment velocity distributions resolved in experiments.
Although in reality, the turbulent forces on moving sediment particles vary in a complex spatio-temporal way, we have approximated the fluid forces on bed load particles as spatially uniform Gaussian white noise. Even though the non-Gaussian aspects of fluid turbulence certainly do impact sediment entrainment \citep{Coleman2019,Celik2014}, this flow model appears more or less justified since sediment transport experiments provide similar velocity distributions regardless of whether the flow is viscous or turbulent \citep{Lajeunesse2010, Charru2004}, and since particle relaxation times are typically long compared to the timescales of turbulent fluctuations \citep{Hofland2006,Schmeeckle2007,Nakagawa1981}.
We modelled particle-bed collision forces as a sequence of instantaneous impulses where the intervals between successive collisions were characterized as exponential random variables. The effect of each collision on the streamwise particle velocities was parameterized by a restitution-like coefficient.
Although such approximate descriptions of particle-particle collisions are common in the theory of granular gases, the setting here is somewhat different than grains in air. Because particles within a viscous flow in general interact at a distance, we should expect the collision model will become poor when the time between subsequent collisions becomes small. Therefore, although the model seems appropriate for saltation, it should be critically examined for ``reptation", when the times between subsequent particle-bed collisions are short. Of course, more realistic flow and collision forces could always be incorporated into Langevin equations for bed load transport, although these might require numerical methods for their solution. The theory of granular gases provides some indication of the types of more nuanced collision models which are possible \citep{Brilliantov2004,other..}.

Sediment transport experiments reveal correlations between particle size and the shape of the bed load velocity distribution.
Experiments with smaller particles tend to give exponential distributions, and those with larger particles give Gaussian distributions.
In fluid dynamics, the dissipation characteristics of particle-particle collisions in viscous flows are known to depend on the particle size and approach velocity through the Stokes number. 
In kinetic theory, it is known that gases of ideal elastic particles generate Gaussian (Boltzmann) velocity distributions, while gases of inelastic grains generate non-Gaussian distributions \citep{Chapman1970,Brilliantov2004}.
Taken together, these ideas suggest that we might relate the shape of the particle velocity distribution to particle size.
We can estimate typical Stokes numbers of colliding bed load particles in experiments as $...$, using with the flow shear velocity and mean streamwise sediment velocity to calculate $V$. Estimating in this way, transport experiments with exponential velocities have $St \sim 1-10$, those with neither exponential nor Gaussian velocities have $St \sim 10-100$, and those with Gaussian velocities have $St > 100$.
In experiments relating restitution coefficient to Stokes number for idealized collisions, restitution coefficients vary sharply from $0$ to $1$ as $St$ ranges from $1$ to $500$ \cite{Marshall2001,Joseph2001,Yang2006}.
Although collision geometry and grain shape certainly complicate the narrative, these values of $St$ are consistent with our model conclusion that the shape of the particle velocity depends on the elasticity of collisions.

@@

\section{Implications for landscape evolution}

Many phenomena of landscape evolution depend on the velocities of individual grains.
The erosion rates of bedrock rivers are for example parameterized by the impact velocities of individual sediment grains against the bedrock surface \citep{Skylar2004,Tingian,Turowski2020}, on which they have a sensitive nonlinear dependence.
It is not much of a stretch to imagine that bedrock erosion could be dominated by the largest impact velocities that originate from the extreme tail of the bedload velocity distribution. Here, we have shown that the weight of this tail depends sensitively on the dissipation characteristics of particle-bed collisions, suggesting that another feedback, that between dissipation and ... may be at play in the evolution of bedrock canyons.

\section{Rivers, hillslopes, and dunes}

It has long been recognized, probably since \citet{Bagnold1941} and certainly today \citep{}, that there are deep analogies between tranpsort phenomena in air and water.
In an ideal perspective, we might imagine writing the governing equations for transport in water as a function of the fluid viscosity, then tuning the viscosity between air and water to obtain a description applying to both spheres.

### 2021-06-29 pref.tex
got some stuff figured out .

\chapter{Preface}

A preface should be written indicating what is to come and what the point of this whole thing is.

>>>>

\chapter{Preface: complementary traditions}

This thesis deals with the transport of coarse sediment in flowing water from the perspective that sediment transport results from the movements of individual grains.
The research has been motivated by the belief that geomorphology as a science will benefit from mechanistic, process-based, mathematical models of sediment transport processes.

Geomorphology has long been characterized by observation, description, and the building of conceptual models, not mathematical ones.
Only relatively recently has the science turned toward quantitative methods, with researchers working to frame observations in terms of underlying processes and to describe them with methods adapted from physics.

Given the complexity of geomorphology problems, a complete reduction of geomorphology to physics would be narrow-minded, but we can nonetheless borrow ways of thinking.
An extremely successful approach in physics has been to construct idealized models with little intention of direct realism, study them deeply to obtain complete understanding, and then, decades or centuries later, to embellish these models with more sophisticated features to describe real-world phenomena.
An example is the block on the spring, the simple harmonic oscillator, which starts as a toy model studied by every first year physics student, but somehow shows up in the deepest inquiries of theoretical physics, from quantum matter to cosmological inflation.
This thesis applies the simple harmonic oscillator approach to some problems in Earth science.
The objective is to build up simple archetypes which can be embellished later, not to build \textbf{the model} of fluvial geomorphology. 
That's for later. (And it's not for me!) CHEERS.
